Hilbert space methods to approximate Gaussian processes using Stan

### 2019-04-29 Paper/manuscript/manuscript_05.tex
some improvements to Section 3

We start with an illustration on how the number of basis functions $m$ and boundary factor $c$ influences the accuracy of the HSGP approximations, separately. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation and the non-linearity of the estimated function, in the sense that fewer basis functions implies inaccurate and smoother effects covariance functions (covariance functions with larger lengthscales) and consequently more linear functional forms, and conversely. 
%:Fewer basis functions imply smoother covariance functions (Paul: when would you consider a covariance function as smooth? Also, I think the main point here is that at long distances the implied covariance function becomes negative which can't be in an exact GP model.) and consequently more linear functional forms. 
Then, the higher the "wigglyness" of the function to be estimated, the more basis functions will be required. 
%(Paul: do we see this in Figure \ref{fig1_Post_J} already?). 
On the other hand, for $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries as well as covariances at long distances.

@@

\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa).
\item There exists a minimum $c$ under which a close approximation will never be achieved regardless of $m$.
\end{itemize}

@@

As stated above, Figure \ref{fig6_lscale_vs_J_vs_c} provides the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c}. If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, it does not exceed it, the approximation may be inaccurate and $m$ should be increased or $c$ decreased (Paul: decreasing $c$ seems dangerous to me if predictions for out of sample data are desired. Shell we mention this? Gabi: We assume that predicting points are already within $S$. Then, a valid $c$ is more related to the lengthscale than to the predicting points. In fact, valid $c$s are restricted in function of the lengthscale: A given lengthscale has a minimum $c$ under which a close approximation will never be achieved regarless $m$ (We stated this in line 502 and also in line 462). For example, a given $c$ can not work for certain lengthscale but for a larger one). We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale provided by Figure \ref{fig6_lscale_vs_J_vs_c}, repeat the process while increasing $m$ or decreasing $c$. Note that, as commented before, $c$ can not be decreased as much as desired because it is restricted to the lengthscale.
%replacing $\ell$ by the newly obtained estimate $\hat{\ell}$.

@@

For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function (Paul: Did we talk about the applied kernels previously in Section 3? Gabi: I did some attempt about that in previous subsection, but I am not sure whether succesfully.) and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both regular GP and HSGP models fitted to those datasets. The lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As commented previously, an accurate estimate of the lengthscale can be a good indicator of a close approximation of the HSGP model to the regular GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the regular GP models as a function of the lengthscale and number of basis functions $m$.

@@

Comparing the information of models approximation provided by Figures \ref{fig8_Tlscale_vs_Elscale} and \ref{fig9_MSE_varing_lscale}, it can be appreciated an agreement between both figures, excepting for small lengthscales and long lengthscales. We observe small RMSE for large lengthscales in Figure \ref{fig9_MSE_varing_lscale}, however there exist mismatch between large lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because big changes on large lengthscales, which might be clearly appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, might result in similar posterior predictions and therefore a small RMSE. On the other side, we observe high RMSE for small lengthscales in Figure \ref{fig9_MSE_varing_lscale}, however there exist match between small lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because very small changes on small lengthscales, which might not be appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, can have big effects on the posteriors predictions and therefore high RMSE.

The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the minimum lengthscales that can be closely approximate in every case, according to Figure \ref{fig6_lscale_vs_J_vs_c}. It can be appreciated a complete agreement between those recommended minimum lengthscales that can be properly fitted in every case following Figure \ref{fig6_lscale_vs_J_vs_c} and the real approximate accuracy measured by the RMSE in Figure \ref{fig9_MSE_varing_lscale}.

>>>>

We start with an illustration on how the number of basis functions $m$ and boundary factor $c$ influences the accuracy of the HSGP approximations, separately. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation and the non-linearity of the estimated function, in the sense that fewer basis functions inaccurately imply larger lengthscales and consequently more linear functional forms.
The higher the "wigglyness" of the function to be estimated, the more basis functions will be required. If $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries as well as covariances at long distances.

@@

\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa).
\item There exists a minimum $c$ below which a close approximation will never be achieved regardless of $m$.
\end{itemize}

@@

As stated above, Figure \ref{fig6_lscale_vs_J_vs_c} provides the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c}. If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, it does not exceed it, the approximation may be inaccurate and $m$ should be increased or $c$ decreased We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale provided by Figure \ref{fig6_lscale_vs_J_vs_c}, repeat the process while increasing $m$ or decreasing $c$. Note that, as commented before, $c$ can not be decreased as much as desired because it is restricted to the lengthscale.
%replacing $\ell$ by the newly obtained estimate $\hat{\ell}$.

@@

For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both regular GP and HSGP models fitted to those datasets. The lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As noted previously, an accurate estimate of the lengthscale can be a good indicator of a close approximation of the HSGP model to the regular GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the regular GP models, as a function of the lengthscale and number of basis functions.

@@

Comparing the accuracy of the lengthscale in Figure 
\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure \ref{fig9_MSE_varing_lscale},
we see that they agree closely with each other for medium lengthscales.
That is, a good estimation of the lengthscale implies a
small RMSE. This is no longer true for very small or large lengthscales.
In small lengthscales, even very small inaccuracies 
may have a strong influence on the posteriors predictions and thus on the RMSE.
In large lengthscales, larger inaccuracies change the posterior predictions
only little and may thus not yield large RMSEs.
The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the 
minimum lengthscale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}. 
We observe that whenever the estimated lengthscale exceeds the minimially
estimable lengthscale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}).
Conversely, when the estimated lengthscale is smaller than the minimally
estimable one, the RMSE becomes very large.

(Paul: Figure 8 and 9 are not rendered correctly in my PDF viewer after 
compiling the tex file, which has something to do with the font of the facet 
and legend labels I believe. Latex also issues warnings about this.)

% Comparing the information of models approximation provided by Figures \ref{fig8_Tlscale_vs_Elscale} and \ref{fig9_MSE_varing_lscale}, it can be appreciated an agreement between both figures, excepting for small lengthscales and long lengthscales. We observe small RMSE for large lengthscales in Figure \ref{fig9_MSE_varing_lscale}, however there exist mismatch between large lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because big changes on large lengthscales, which might be clearly appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, might result in similar posterior predictions and therefore a small RMSE. On the other side, we observe high RMSE for small lengthscales in Figure \ref{fig9_MSE_varing_lscale}, however there exist match between small lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because very small changes on small lengthscales, which might not be appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, can have big effects on the posteriors predictions and therefore high RMSE.

%The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the minimum lengthscales that can be closely approximate in every case, according to Figure \ref{fig6_lscale_vs_J_vs_c}. It can be appreciated a complete agreement between those recommended minimum lengthscales that can be properly fitted in every case following Figure \ref{fig6_lscale_vs_J_vs_c} and the real approximate accuracy measured by the RMSE in Figure \ref{fig9_MSE_varing_lscale}.
